The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-09-30 article.tex
Some more text fixes/modifications. Added todo notes

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collision detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a difference in output
  between  
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

The API has hardly changed, but the internals have significantly. Besides
the aforementioned overlap resolution, the engine also handles large ring
systems much better,
and handles stereochemistry of the input. Previously, double bond stereochemistry
was supported, but the new engine more efficiently uses this when creating 2D
layouts. Furthermore, the engine assigns wedge bond information based on
tetrahedral stereochemistry. These features are exemplified by the following
code and the resulting layout depiction in Figure~\ref{fig:sdgstereo}.

@@

  A chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute the composition
of a compound without describing how atoms are bonded. With the rise of
metabolomics it has become increasingly relevant to have full support for these
in cheminformatics libraries~\cite{RojasCherto2011}.

@@

  The SMILES parsing has been replaced code in the external Beam project~\cite{Beam}.
  This BSD-licensed SMILES parser is a complete implementation of the SMILES
  and OpenSMILES (\url{http://opensmiles.org/}) specifications by one of the
  authors (including stereochemistry), and is independent of
  the CDK library. The SmilesParser API uses this library underneath, and the
  Beam API is hidden by this class. The most significant change here is that
  the SMILES parser now automatically locates the positions of double bonds,
  but this can be turned off:

@@

  Substructure matching is fundamental to many other procedures. Since CDK 1.2 functionality has been added to handle the SMARTS query language. The SMARTS language is supported well including: Stereochemistry, Component Grouping, Atom Maps (to match reaction transformations). A new \textit{Pattern} API has been added to CDK \cdkversion{}, the API simplifies finding, filtering, and transform results. The API is immutable allowing a pattern to be initialized once and then matched against several molecules or reactions across multiple threads, during initialization the pattern is inspected as to determine what invariants will be needed (e.g. ring size) and only calculating what is needed. The internal matching algorithms provide a lazy iterator, the lazy evaluation means the next match is only calculated when it is needed. The API handles reactions in addition to molecules, both can be specified as either queries or targets.

@@

  CDK \cdkversion{} includes large improvements to algorithm efficiency, an isolated benchmark demonstrates relative improvements between recent CDK versions and to other open source toolkits (Table~5\ref{tab:smigrep}). The efficiency improvements are a combination of optimising data structures and key molecule processing algorithms (e.g. Kekulisation and Aromaticity) needed before a SMARTS match can be run.

@@

  Ring finding is a key functionality in a cheminformatics library, and the CDK
  knows a long history of ring finding~\cite{May2014}. Particularly, non-redundant
  ring sets have seen particular interest, such as the smallest set of smallest
  rings, for which the CDK implements two classical algorithms~\cite{Figueras1996,Berger2004}.
  Recent work by one the authors has implemented a new, faster algorithm, allowing
  searching for various types of (non-redundant) ring sets~\cite{May2014}. These
  are available via the new Cycles API:

@@

  Aromaticity has seen many definitions in the past and for cheminformatics it
  frequently is algorithmically defined. The outcome of an aromaticity calculation
  depends on a number of atom type features and heuristics, which in literature
  often are not well defined. Based on this information several different
  algorithmic definitions of aromaticity can be defined. Older CDK versions had
  various models implemented and the code was scattered throughout the library.
  This was unified resulting in three models, of which two are based on the
  CDK atom typer, to provide the number of electrons donated by each atom to
  the delocalized $\pi$ system. The difference between these two models is
  how double bonds pointing outwards from a ring are contributing.

  The current CDK version further generalizes the idea that aromaticity is a
  model, and provides an API that allows you to select out of various options
  providing greater interoperability with other toolkits. The new
  Aromaticity class allows to build a custom model by selecting and combining
  options. For example, to reproduce the functionality of the previous
  CDKAromaticity class:

@@

Here, the CDK model for counting donated electrons is used, as the ring systems
the CDK previously included in aromaticity counting, which was limited in the 
number of fused rings systems. However, an alternative aromaticity calculator
that does consider all possible ring systems can now be easily created with:

@@

The molfile format is still very popular, despite it being a proprietary
format it has become a de facto standard. The format forms the core of the larger
CTfile family which was originally developed by MDL Information Systems~\cite{Dalby92}. The
current format specification is published by BIOVIA and available on 
request~\cite{ctfilespec}.

@@

Originally, the CDK was developed as a shared library between
JChemPaint~\cite{krause2000jchempaint} and Jmol~\cite{Willighagen2007jmol,Hanson2010}.
JChemPaint used a MVC approach with an event-passing mechanism to update the view when the model was
changed. This can cause an cascade of change events being passed around. To address,
interfaces were introduced allowing multiple implementations of the core interfaces.
% TODO: check if the interfaces have previously been defined
The IChemObjectBuilders play an important role here, allowing implementations of the
interfaces to be instantiated without the need of explicitly referencing those implementations.

However, the CDK 1.0 and 1.2 implementations of the IChemObjectBuilder had one method for
each constructor, resulting in very large interface. Moreover, the API changed whenever
a new class was introduced, and existing methods changed when constructors were updated.
To simplify the API, the new IChemObjectBuilder collapsed all methods returning new
implementations into a single method, which takes as a first parameter the class of the
interface that is wished to be constructed. All further parameters are passes as
parameter to the class constructor.

For example, to construct a new atom from its element symbol, one would now write:

@@

Increasingly, the CDK library is now written against the interfaces, and when new instanced
are needed, these builders are being used. This allows to run a certain CDK-based
application with a specific builder, aimed at a particular use case. The original
implementation is available via the DefaultChemObjectBuilder, which creates
classes that pass around change events, just like the original CDK data classes.
However, a SilentChemObjectBuilder has been introduced that does not pass around
change events, which makes code run about 10-20\% faster.
The third builder is the DataDebugChemObjectBuilder which generates debug information
for all changes to the content of the data classes. This can be useful for
debugging and other forms of code inspection.

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~3\ref{tab:fingerprints}.

@@

As the library grew over the years, so did the complexity of the maintenance:
Increasingly frequent, the main branch failed to compile, and bug fixing become
more difficult with unexpected side effects. Often fixing a bug in one part of
the code, broke some other code which made the incorrect assumptions about the
fixed code. Such is inevitable when the number of developers increases.

To address these issues, we have adopted a number of coding standards. By no
means these are meant to implement the best practices of source code development,
which is not practically unfeasible with limited resources; instead, they attempt
to find a balance between increasing code maintainability and being flexible
enough to allow efficient code development. However, we appreciate the subjective 
nature of this statement, and some adopted guidelines have been heavily discussed
and debated in the CDK community.

Arguably, perhaps the biggest factor in improved code quality is that we
instantiated a peer review process where any functionality changing patch is
required to be reviewed by one independent, senior CDK developer for the development
branch, and two reviewers for stable branches. This patch development system
is supported by a number of automated validations steps as outlined below.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~\ref{fig:deps}.

@@

  The quality of JavaDoc of the CDK was originally tested with DocCheck, and
  later replaced by a custom written tool called OpenJavaDocCheck. With the move
  to Maven (explained elsewhere) which does not have integration for this tool,
  we adopted CheckStyle (\url{http://checkstyle.sourceforge.net/}). This tool
  reports about missing documentation and on documentation which is not properly
  annotated in the Java source files.

@@

Another change made over the past years is the move to a Git-based version
control system. Advantages the projects makes advantage of is the distributed
nature of Git repositories, and easier branching and making available of
patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced SourceForge
as the main source code hosting service
where we can use novel approaches for commenting on code (peer review),
pull requests, etc.
These new features simplify our code review process.

@@

Another recent change is the choice of build system: we have moved away from
Ant to Maven to deal with dependencies and compile and test the library.
To keep the original idea of CDK modules, this move required to split up the
source code in multiple source folders. Fortunately, modern integrated
development environments have no trouble handling this, removing the original
argument to have everything in one source folder.

On the other hand, for many modules, the test code is now more closely
linked to the code being tested: both reside in the same folder, though we
adhere to the Maven custom to have src/main/java and a src/test/java folders.
For a few modules, however, this solution introduces circular dependencies, in
which case a separate Maven module is created for the tests.

@@

Since the last CDK publication in 2006, the library has been improved in many
aspects. The functionality has seen numerous additions and rewrites, making the
library more functional and a lot faster than before. Updates on the common
SMILES and molfile formats and the improved structure diagram generation are
very visible and benefit many of the tools using the CDK.
Furthermore, the stability of the development model has significantly improved,
providing greater stability of the library over time. The project adopted peer
review and quality control.

>>>>

%% Comment out before submission 
\usepackage[colorinlistoftodos,textsize=small]{todonotes}

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. In particular, collision detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a difference in output
  between  
  the old code base, with and without overlap resolving, and with the
  new code. The latest implementation simplifies the generation of 2D
  coordinates as shown below:

@@

  While the API itself has not been significantly changed, the
  internals have been revamped. In addition to improved overlap
  resolution noted above, the engine appropriately handles large ring
  systems and maintains input stereochemistry. While previous CDK
versions supported 
  double bond stereochemistry the new engine is more efficient in
  using this information when generate 2D layouts. Furthermore, the
  engine assigns wedge bond information based on tetrahedral
  stereochemistry. These features are exemplified by the following
  code and the resulting layout depiction in
  Figure~\ref{fig:sdgstereo}.

@@

A chemical formula is the simplest chemical representation of a
compound.  It defines the number of isotopes or elements that
compose  a compound without describing how atoms
are bonded. With the rise of metabolomics it has become increasingly
relevant to have full support for these in cheminformatics
libraries~\cite{RojasCherto2011}.

@@

The SMILES parsing has been replaced by code from the external Beam project~\cite{Beam}.
This BSD-licensed SMILES parser is a complete implementation of the SMILES
and OpenSMILES (\url{http://opensmiles.org/}) specifications by one of the
authors (including stereochemistry), and is independent of
the CDK library. The SmilesParser API uses this library underneath, and the
Beam API is hidden by this class. The most significant change here is that
the SMILES parser now automatically locates the positions of double bonds,
but this can be turned off:

@@

Substructure matching is fundamental cheminformatics operation and
plays a key role in many other functions such as fingerprint,
descriptor generation and atom typing. Since CDK 1.2, functionality
has been added to handle the SMARTS query language. The SMARTS
language is supported well including advanced features such as
stereochemistry, component grouping, and atom
maps (to match reaction transformations). A new \textit{Pattern} API
has been added to CDK \cdkversion{}, which simplifies finding,
filtering, and transform results. The API is immutable allowing a
pattern to be initialized once and then matched against several
molecules or reactions across multiple threads. During initialization
the pattern is inspected as to determine what invariants will be
needed (e.g. ring size) and only calculating what is needed. The
internal matching algorithms provide a lazy iterator, such that the
next match is only computed when it is needed. The API handles
reactions in addition to molecules, and both can be specified as
either queries or targets.

@@

CDK \cdkversion{} includes large improvements to algorithm
efficiency. A simple benchmark experiment demonstrates the relative
improvements between recent CDK versions and other open source
toolkits (Table~5\ref{tab:smigrep}). The efficiency improvements are a
combination of optimising data structures and key molecule processing
algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
match can be run.

@@

Ring finding is a key functionality in a cheminformatics library, and
the CDK knows a long history of ring finding~\cite{May2014}. In
particular, non-redundant ring sets have seen particular interest,
such as the smallest set of smallest rings, for which the CDK
implements two classical algorithms~\cite{Figueras1996,Berger2004}.
Recent work has implemented a new, faster algorithm, allowing
searching for various types of (non-redundant) ring
sets~\cite{May2014}. These are available via the new Cycles API:

@@

Aromaticity has seen many definitions in the past and for
cheminformatics it frequently is algorithmically defined. The outcome
of an aromaticity calculation depends on a number of atom type
features and heuristics, which are often ambiguously defined in the
published literature. Based on this information several different
algorithmic definitions of aromaticity can be defined. Older CDK
versions had various models implemented and the code was scattered
throughout the library, resulting in an inconsistent approach to
calling methods to compute aromaticity and a significant maintenance
burden.  This was unified in the current version, resulting in three
models, of which two are based on the CDK atom typer. The difference
between these two models is how contributions from exocyclic double
bonds are handled.

The current CDK version further generalizes the idea that aromaticity
is a model, and provides an API that allows the user to select one of
several aromaticity models, leading to greater interoperability with
other toolkits. The new \texttt{Aromaticity} class allows to build a
custom model by selecting and combining options. For example, to
reproduce the functionality of the previous \texttt{CDKAromaticity} class:

@@

Here, the CDK model for counting donated electrons is used, along with
the rings systems that were identified by the older algorithm in
previous versions that was limited in the number of fused rings
systems that were considered. However, an alternative aromaticity
calculator that considers all possible ring systems can now be
easily created with:

@@

The molfile format is still very popular and despite it being a proprietary
format it has become a de facto standard. The format forms the core of the larger
CTfile family which was originally developed by MDL Information Systems~\cite{Dalby92}. The
current format specification is published by BIOVIA and available on 
request~\cite{ctfilespec}.

@@

Originally, the CDK was developed as a shared library between
JChemPaint~\cite{krause2000jchempaint} and
Jmol~\cite{Willighagen2007jmol,Hanson2010}.  JChemPaint used a MVC
approach with an event-passing mechanism to update the view when the
model was changed. This can cause an cascade of change events being
passed around. This was not always a desirable feature, especially for
non-UI code. To address this, interfaces were introduced allowing
multiple implementations of the core interfaces.
% TODO: check if the interfaces have previously been defined
The \texttt{IChemObjectBuilders} play an important role here, allowing
implementations of the interfaces to be instantiated without the need
of explicitly referencing those implementations.\todo{The motivation
  for IChemObjectBuilders is not well defined.}

However, the CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had one method for
each constructor, resulting in a very large interface. Moreover, the API changed whenever
a new class was introduced, and existing methods changed when constructors were updated.
To simplify the API, the new \texttt{IChemObjectBuilder} collapsed all methods returning new
implementations into a single method, which takes as a first parameter the class of the
interface that is wished to be constructed. All further parameters are passed as
parameters to the class constructor.

For example, to construct a new atom from its element symbol, one
would now write:\todo{might be good to display code using older version}

@@

The CDK library is now written against the interfaces\todo{this may
  not mean to a non-expert Java coder}, and when new instances are
needed, these builders are used. This allows to run a certain
CDK-based application with a specific builder, aimed at a particular
use case. The original implementation is available via the
\texttt{DefaultChemObjectBuilder}, which creates classes that pass
around change events, as in the older CDK versions.  However, a
\texttt{SilentChemObjectBuilder} is available, that does not support
change events, resulting in speedups of 10\%-20\%.  The third builder
is the \texttt{DataDebugChemObjectBuilder} which generates debug
information for all changes to the content of the data classes. This
can be useful for debugging and other forms of code inspection.

\subsubsection*{Molecular fingerprints}
Molecular fingerprints are another area that seen significant
development. Traditionally fingerprints were represented using the
\texttt{BitSet} class from the Java library. While using this class
allowed the use of pre-existing methods to manipulate bit strings, it
keeps a vector of bits in memory. The solution was excellent for
hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a
fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
indexing space using this approach would be highly
memory-inefficient. To allow for multiple fingerprint representation
implementations a bit fingerprint interface was introduced\todo{is
  there any more information about this?}. Also,
although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer
vectors supported in CDK as well. The fingerprints currently existing
in CDK are listed in Table~3\ref{tab:fingerprints}.

@@

As the library grew over the years, so did the complexity of the
maintenance. The main branch frequently failed to compile and bug
fixes became more onerous due to unexpected side effects.  Often
fixing a bug in one part of the code, broke some other code which made
the incorrect assumptions about the fixed code. With the increase in
the CDK developer community, such issues were inevitable, in the
absence of any formal coding and testing standards.

To address these issues, we have adopted a number of coding
standards. While not a comprehensive implementation of software
engineering best practices, they attempt to find a balance between
increasing code maintainability and being flexible enough to allow
efficient code development. However, we appreciate the subjective
nature of this statement, and some adopted guidelines have been
heavily discussed and debated in the CDK community.

Arguably, perhaps the biggest factor in improved code quality is a
peer review process where any functionality changing patch is required
to be reviewed by one independent, senior CDK developer for the
development branch, and two reviewers for stable branches. This patch
development system is supported by a number of automated validations
steps as outlined below.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
between the CDK modules are depicted in
Figure~\ref{fig:deps}.\todo{the dependency graph is not very legible}

@@

The quality of the JavaDocs was originally tested with DocCheck, and
later replaced by a custom written tool called OpenJavaDocCheck. With the move
to Maven (explained elsewhere) which does not have integration for this tool,
we adopted CheckStyle (\url{http://checkstyle.sourceforge.net/}). This tool
reports on missing documentation and on documentation which is not properly
annotated in the Java source files.

@@

Older versions of the CDK employed Subversion for version control. A
few years back, the project switched to the Git version control
system. A key advantage of this shift is the ability to have
distributed repositories, easier branching and provision for
patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced
SourceForge as the main source code hosting service where we can use
novel approaches for commenting on code (peer review), pull requests,
etc.  These new features simplify our code review process.

@@

The build system has been converted from Ant to Maven. The shift was
motivated by the easier dependency handling, cleaner separation of
testing code from the main library and automated packaging. The move
to modules necessitated splitting the original monolithic source code
tree in to per-module source folders. While this makes the on-disk
layout of the source code more complex, this is usually hidden by
modern IDEs.

As a result for many modules, the test code is now more closely linked
to the code being tested: both reside in the same folder, though we
adhere to the Maven custom to have \texttt{src/main/java} and a
\texttt{src/test/java} folders.  For a few modules, however, this
solution introduces circular dependencies, in which case a separate
Maven module is created for the tests.

@@

Since the last CDK publication in 2006, the library has been improved
in many aspects including architecture, new functionality and improved
code testing, management and deployment. These changes have led a more
functionally rich cheminformatics library, with significant
performance improvements. Updates on the common SMILES and molfile
formats and the improved structure diagram generation are very visible
and benefit many of the tools using the CDK.  Furthermore, the
stability of the development model has significantly improved,
providing greater stability of the library over time. The project
adopted peer review and quality control.
